---
title: "Getting Started with insitu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GettingStarted}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r docsetup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction
Extant biodiversity and species distribution across the globe result from the interplay of three fundamental processes: speciation, extinction, and dispersal. Evolutionary biologists often work to disentangle the contributions of each of these processes to the diversity we observe today. Furthermore, associating these processes with the geography on which they occur provides researchers with a clearer picture of how extant biodiversity emerged.

We can associate speciation with geography by considering *in situ* speciation events. A speciation event is considered to be *in situ* when it occurs within a specific geographical range, as opposed to a lineage colonizing a location from elsewhere. The importance of *in situ* speciation for understanding diversity patterns has been emphasized by decades of work, much of which discusses how *in situ* speciation contributes to species richness on islands. 

The `insitu` R package provides a suite of functions that allows the user to estimate *in situ* speciation events and quantify their contribution to biodiversity in discrete geographical units, with a primary focus on island systems and island-like environments. This vignette will cover the functions from the `insitu` package that are necessary to perform a basic analysis that classifies nodes on a phylogeny as *in situ* speciation, colonization, or dispersal events, as well as computing diversity decomposition metrics and visualizing results.

This vignette will walk through the following steps in the `insitu` pipeline:

* Data Preparation
* Classifying Events and Diversity Decomposition
* Indices and Support
* Adaptive Radiation Bursts


## Data Preparation
For its most basic functions, the `insitu` package requires the user to input a phylogeny and a list of locations that each species of interest occupies. The phylogenetic tree should be ultrametric and include all species listed in the location data, but if there are any mismatches in these datasets, the `insitu::prep_phylo()` and `insitu::match_island_phylo()` functions will address them. The example data here includes 30 island-dwelling *Anolis* lizard species sampled from the phylogenetic tree of all *Anolis* species included in Patton et al. (2021).

```{r dataprep}
# Load insitu library
library(insitu)

# Read in example tree
tree <- ape::read.tree(system.file("extdata/Patton_etal_trimmed.tree",
                                       package = "insitu"))

# Handle non-ultrametricity and incomplete taxon sampling
tree <- prep_phylo(tree)

# Read in example dataset
dat <- read.csv(system.file("extdata/anolis_dat.csv",
                            package = "insitu"))

# Ensure that species represented in both the tree and data are the same
matched <- match_island_phylo(phy = tree, locs = dat)
```
The `insitu::match_island_phylo()` function removes species that are not included in both the tree and the data, and it prints out the names of species removed from each. It also prints out the first five locations in the dataset as a way for users to determine whether the correct locations will be included in downstream analyses.

## Classifying Events
The `insitu` package provides multiple functions for estimating ancestral character states and classifying nodes on a phylogeny as *in situ* speciation, colonization, or dispersal events. First, we will use different methodology to estimate ancestral character states via ancestral state reconstruction (ASR). Then, we will classify each node as an appropriate event based on the estimated history of that node. Finally, we will estimate and visualize diversity decomposition at each node.

## Geography-Based Ancestral State Reconstruction

```{r geo_asr}
# Use the tree and PAM from the "matched" object to run an ancestral state reconstruction
asr_insitu <- run_geo_asr(phy = matched$phy, PAM = matched$PAM)

```

## Classify Speciation Events

```{r classify}
# Using the ASR from above, determine whether speciation events occurred in situ
map_events <- map_insitu_events(recons = asr_insitu,
                                phy = matched$phy,
                                PAM = matched$PAM,
                                threshold = 0.5)

# Look at the first 5 entries in the event data frame
head(map_events, n = 5)

```

## Visualize Speciation Events on the Phylogeny

```{r plot_phylo, fig.width = 8, fig.height = 6}

plot_classified_phylo(phy = matched$phy, events = map_events)

```

## Partition Species Richness Into Different Components (in situ speciation, import, and export events)

```{r decompose}
div <- decompose_diversity(phy = matched$phy,
                           events = map_events,
                           PAM = matched$PAM)

# Look at the first 5 rows of the classification data frame
head(div, n = 5)
```

## Visualize Diversity Decomposition

```{r plot_div, fig.width = 7, fig.height = 5}
# Visualize diversity decomposition per island in a stacked bar chart
plot_diversity_decomposition(div)

# Determine whether:
# - In situ speciation, import, and export counts differ across islands
# - In situ and colonization rates differ across islands
compare_islands(div)
```

## Determine Colonization and *in situ* Speciation Rates
```{r rates}
# Colonization rate (how many times each species has colonized an island)
colonization_rate(phy = matched$phy, events = map_events)

# in situ speciation rate (how many times each species is associated
#                          with an in situ speciation event)
insitu_speciation_rate(phy = matched$phy, events = map_events)

```

## Calculate the in situ Speciation Index
```{r index}
# Per-siland in situ speciation index
insitu_speciation_index(phy = matched$phy, events = map_events)
```

## Determine Node Support
```{r confidence}
# Determine per-node support for in situ speciation events
insitu_confidence(recons = asr_insitu)
```

## Detect Adaptive Radiation Bursts
```{r bursts1, fig.width = 8, fig.height = 6}
# Identify nodes associated with radiation bursts
detect_radiation_bursts(phy = matched$phy,
                        events = map_events,
                        time_window = 5,
                        min_events = 2,
                        plot = TRUE)

```

## Classify Nodes, But Without Using ASR First
```{r classification, fig.width = 8, fig.height = 6}
# Identify nodes as within-island, between-island, or ambiguous
# Classification is only based on tip composition
node_classification <- identify_island_nodes(phy = matched$phy, PAM = matched$PAM)
# Add colors to nodes based on the classification determined above
plot_island_nodes(phy = matched$phy, node_class = node_classification, PAM = matched$PAM)

```

## Literature Cited
* Patton, A. H., Harmon, L. J., del Rosario Castañeda, M., Frank, H. K., Donihue, C. M., Herrel, A., & Losos, J. B. (2021). When adaptive radiations collide: Different evolutionary trajectories between and within island and mainland lizard clades. Proceedings of the National Academy of Sciences, 118(42), e2024451118
