---
title: "Getting started with insitu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with insitu}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

The `insitu` package classifies ancestral nodes on a dated phylogeny as
in-situ speciation, colonization, or dispersal events using distributional
and phylogenetic data. It then provides functions to partition species
richness into its historical components, compute island-specific indices,
and visualize the results.

This vignette covers the core workflow from data preparation through
event classification, diversity decomposition, and island ontogeny
correction, using 30 *Anolis* lizard species from the Caribbean Islands
as a worked example.

## Data preparation

The pipeline requires a time-calibrated phylogeny and a table of species
occurrences with at least two columns: `species` and `locale`. If a species
occurs on the mainland, record its locality as `"Mainland"`. Species must
appear in both the tree and the occurrence table.

```{r dataprep}
library(insitu)

tree <- ape::read.tree(
  system.file("extdata/Patton_etal_trimmed.tree", package = "insitu")
)
dat  <- read.csv(
  system.file("extdata/anolis_dat.csv", package = "insitu")
)
```

`prep_phylo` handles non-ultrametricity and stores a sampling fraction
correction that downstream rate functions use automatically. Most empirical
trees have small rounding errors that make them non-ultrametric; this step
fixes those before anything else runs.

```{r prep}
tree <- prep_phylo(tree, force_ultrametric = TRUE, sampling_fraction = 1)
```

`match_island_phylo` prunes the tree to species in the occurrence data,
builds the presence-absence matrix (PAM), and prints a summary of what
was retained and what was dropped. Set `exclude = FALSE` to keep mainland
outgroup species in the tree for root anchoring.

```{r match}
matched <- match_island_phylo(phy = tree, locs = dat)
```

The `matched` object is a list with three elements: `$phy` (the pruned
tree), `$PAM` (the presence-absence matrix), and `$sp_df` (a per-island
species summary). Everything downstream takes `matched$phy` and
`matched$PAM` as inputs.

## Ancestral state reconstruction

`run_geo_asr` runs a discrete-state ancestral state reconstruction for
each island in the PAM using `ape::ace`. Each island is treated as a
distinct state alongside the mainland, so the reconstruction estimates
the probability that each ancestral node was present on each specific
island. The transition model `"ER"` assumes equal rates between all
state transitions and is the recommended starting point for most
analyses.

```{r asr, warning = FALSE}
asr_insitu <- run_geo_asr(phy = matched$phy, PAM = matched$PAM, model = "ER")
```

The result is a named list of `ace` objects, one per island. The
`StochasticCharMapping` vignette covers `simmap_insitu` as an alternative
that propagates ASR uncertainty through stochastic character mapping.

## Event classification

`map_insitu_events` classifies each internal node and tip based on the
ancestral reconstruction. A node is in-situ when both it and its parent
reconstruct as the same island above the probability threshold. When the
parent was on a different island the event is inter-island, and when the
parent was on the mainland the event is a colonization. The default
threshold of 0.5 is a reasonable starting point; the `Simulating island
radiations` vignette shows how to calibrate the threshold for your own
system.

```{r classify}
map_events <- map_insitu_events(
  recons    = asr_insitu,
  phy       = matched$phy,
  PAM       = matched$PAM,
  threshold = 0.5
)

head(map_events, n = 5)
```

Each row corresponds to one node-island combination. The `in_situ` column
is `TRUE` when in-situ speciation was inferred. The `export` and `import`
columns record the source and destination island for each transition,
preserving the directionality of dispersal events for downstream analysis.

## Visualization

`plot_classified_phylo` plots the phylogeny with colored nodes at
in-situ speciation events and pie charts at the tips showing each
species' island distribution. Colors follow the Okabe-Ito palette and
are consistent across all plotting functions.

```{r plot_phylo, fig.width = 9, fig.height = 6}
plot_classified_phylo(phy = matched$phy, events = map_events, PAM = matched$PAM)
```

## Diversity decomposition

`decompose_diversity` counts in-situ speciation, import, and export events
per island and computes the proportion of each island's richness that
derives from in-situ speciation. Import events are cases where a lineage
arrived on an island from elsewhere; export events are cases where a
lineage left.

```{r decompose}
div <- decompose_diversity(
  phy    = matched$phy,
  events = map_events,
  PAM    = matched$PAM
)

div
```

`plot_diversity_decomposition` shows these counts as a stacked bar chart,
making it easy to compare the relative contributions of in-situ speciation,
immigration, and emigration across islands at a glance.

```{r plot_div, fig.width = 7, fig.height = 5}
plot_diversity_decomposition(div)
```

## In-situ speciation index

The in-situ speciation contribution index (ISCI) standardizes in-situ
event counts by the total branch length of each island's subtree, allowing
comparisons across islands of different sizes and ages. A higher ISCI means
more in-situ speciation relative to the evolutionary time represented on
that island.

```{r index}
insitu_speciation_index(phy = matched$phy, events = map_events)
```

## Node support

`insitu_confidence` extracts the per-node posterior probability of island
presence for every internal node and every island, binning each into low
(< 0.5), moderate (0.5--0.75), or high (>= 0.75) support. This is useful
for sensitivity analyses and for deciding whether to raise or lower the
threshold in `map_insitu_events`.

```{r confidence}
conf <- insitu_confidence(recons = asr_insitu)
head(conf, n = 15)
```

## Island ontogeny

A node classified as in-situ on an island that did not exist at the time
of that node is biologically impossible. `age_correct_events` compares
each node's age to the formation age of the island it is assigned to
and flags events that predate the island.

Island ages for the Caribbean are approximate; the values below are
illustrative of the workflow.

```{r ontogeny}
ages <- c(
  Cuba             = 42,
  Hispaniola       = 50,
  Jamaica          = 25,
  "Puerto Rico"    = 46,
  "Cat Island"     = 10,
  "Ile de la Tortue" = 15
)

events_corrected <- age_correct_events(
  events           = map_events,
  phy              = matched$phy,
  island_ages      = ages,
  remove_predating = FALSE
)

# Which events predate their island?
events_corrected[events_corrected$predates_island == TRUE,
                 c("node", "island", "node_age", "island_age")]
```

Setting `remove_predating = TRUE` removes those events and returns a
cleaned table that can be passed directly to `plot_classified_phylo` or
`decompose_diversity`.

```{r ontogeny_clean, fig.width = 9, fig.height = 6}
events_clean <- age_correct_events(
  events           = map_events,
  phy              = matched$phy,
  island_ages      = ages,
  remove_predating = TRUE
)

plot_classified_phylo(phy = matched$phy, events = events_clean, PAM = matched$PAM)
```

`insitu_rate_by_island_age` computes the in-situ rate using island age as
the time denominator rather than total path length, which is the correct
denominator when comparing diversification dynamics across islands of
different geological ages. This function requires the output of
`age_correct_events` as input.

```{r age_rate}
insitu_rate_by_island_age(
  phy         = matched$phy,
  events      = events_corrected,
  island_ages = ages
)
```

## Literature cited

Patton AH, Harmon LJ, del Rosario Castañeda M, Frank HK, Donihue CM,
Herrel A, Losos JB (2021) When adaptive radiations collide: Different
evolutionary trajectories between and within island and mainland lizard
clades. Proceedings of the National Academy of Sciences 118:e2024451118.

Whittaker RJ, Triantis KA, Ladie RJ (2008) A general dynamic theory of
oceanic island biogeography. Journal of Biogeography 35:977-994.
