---
title: "Getting Started with insitu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GettingStarted}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r docsetup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction
Extant biodiversity and species distribution across the globe result from the interplay of three fundamental processes: speciation, extinction, and dispersal. Evolutionary biologists often work to disentangle the contributions of each of these processes to the diversity we observe today. Furthermore, associating these processes with the geography on which they occur provides researchers with a clearer picture of how extant biodiversity emerged.

We can associate speciation with geography by considering *in situ* speciation events. A speciation event is considered to be *in situ* when it occurs within a specific geographical range, as opposed to a lineage colonizing a location from elsewhere. The importance of *in situ* speciation for understanding diversity patterns has been emphasized by decades of work, much of which discusses how *in situ* speciation contributes to species richness on islands. 

The `insitu` R package provides a suite of functions that allows the user to estimate *in situ* speciation events and quantify their contribution to biodiversity in discrete geographical units, with a primary focus on island systems and island-like environments. This vignette will cover the functions from the `insitu` package that are necessary to perform a basic analysis that classifies nodes on a phylogeny as *in situ* speciation, colonization, or dispersal events, as well as computing diversity decomposition metrics and visualizing results.

This vignette will walk through the following steps in the `insitu` pipeline:

* Data Preparation
* Classifying Events and Diversity Decomposition
* Rates, Indices, and Support
* Adaptive Radiation Bursts


## Data Preparation
For its most basic functions, the `insitu` package requires the user to input a phylogeny and a list of locations that each species of interest occupies. The phylogenetic tree should be ultrametric and include all species listed in the location data, but if there are any mismatches in these datasets, the `insitu::prep_phylo()` and `insitu::match_island_phylo()` functions will address them. The example data here includes 30 island-dwelling *Anolis* lizard species sampled from the phylogenetic tree of all *Anolis* species included in Patton et al. (2021).

```{r dataprep}
# Load insitu library
library(insitu)

# Read in example tree
tree <- ape::read.tree(system.file("extdata/Patton_etal_trimmed.tree",
                                       package = "insitu"))

# Handle non-ultrametricity and incomplete taxon sampling
tree <- prep_phylo(tree)

# Read in example dataset
dat <- read.csv(system.file("extdata/anolis_dat.csv",
                            package = "insitu"))

# Ensure that species represented in both the tree and data are the same
matched <- match_island_phylo(phy = tree, locs = dat)
```

The `insitu::match_island_phylo()` function removes species that are not included in both the tree and the data, and it prints out the names of species removed from each. It also prints out the first five locations in the dataset as a way for users to determine whether the correct locations will be included in downstream analyses. The `matched` object is a list of three items: a presence-absence matrix (PAM) that summarizes the islands each species occupies (1 for presence and 0 for absence), a tibble that lists each island and all species that are present on each island, and a phylogenetic tree with tips matching those in the user-provided dataset.

## Classifying Events and Diversity Decomposition
The `insitu` package provides multiple functions for estimating ancestral character states and classifying nodes on a phylogeny as *in situ* speciation, colonization, or dispersal events. First, we will estimate ancestral character states via ancestral state reconstruction (ASR). Then, we will classify each node as an appropriate event based on the estimated history of that node. Next, we will estimate and visualize diversity decomposition at each node. Finally, we will cover methodology for classifying events without ancestral state reconstruction.

### Geography-Based Ancestral State Reconstruction
The first method for estimating ancestral states uses the PAM included in the `matched` object created above to identify extant geographical character states for each species. Then, ancestral character states are estimated across the provided phylogeny for each island in the dataset.

```{r geo_asr, warnings = FALSE}
# Use the tree and PAM from the "matched" object to run an ancestral state reconstruction
asr_insitu <- run_geo_asr(phy = matched$phy, PAM = matched$PAM)

```

The `asr_insitu` object is a list of `ace` objects describing the ASR relevant to each island in the dataset. The number of elements in the list resulting from `insitu::run_geo_asr()` depends on the number of islands. In this case, there are six elements in the `asr_insitu` list.

The `insitu` package also provides functionality for using stochastic character mapping to estimate ancestral states. The workflow for this methodology is outlined in the `Stochastic Character Mapping With insitu` vignette. At the end of the `Classifying Events and Diversity Decomposition` section, we also cover methods for classifying nodes that do not require ancestral state reconstruction.

### Classify Speciation Events
Now that we have estimated ancestral character states, we can use them to determine whether speciation events occurred *in situ* (within a given island) by determining whether the most recent common ancestor between two tips is likely to have occurred on the same island as the descendants. For this task, we will use the `insitu::map_insitu_events()` function. This function requires the user to input the ancestral state reconstructions returned by `insitu::run_geo_asr()`, the matched phylogeny and PAM returned by `insitu::match_island_phylo()`, and a threshold probability for determining whether *in situ* speciation is evident. The default threshold probability is 0.5, meaning that if the probability that an ancestor occupied a given island is 0.5 or greater, and the descendants of that node also have the same island state, the `insitu::match_island_phylo()` function will determine that an *in situ* speciation event occurred.

```{r classify}
# Using the ASR from above, determine whether speciation events occurred in situ
map_events <- map_insitu_events(recons = asr_insitu,
                                phy = matched$phy,
                                PAM = matched$PAM,
                                threshold = 0.5)

# Look at the first 5 entries in the event data frame
head(map_events, n = 5)

```

The resulting data frame includes: the node number, the island state, the probability that the ancestor of a given node shares the same island state, and whether this node represents an *in situ* speciation event. The bottom of the data frame includes information on estimated transitions between islands, their associated nodes, and the islands associated with the transition. An island in the "export" column were originally occupied by the species that now occupy the island in the "import" column.

### Visualize *in situ* Speciation Events on the Phylogeny
We can visualize the probabilities and classifications provided by the `insitu::map_insitu_events()` function on the *Anolis* phylogeny using the `insitu::plot_classified_phylo()` function. When the ancestor presence probability has reached or exceeded the treshold (0.5 in this case), a colored dot will be placed on the associated node that corresponds to the island on which the *in situ* speciation event occurred. 

```{r plot_phylo, fig.width = 8, fig.height = 6}

plot_classified_phylo(phy = matched$phy, events = map_events, PAM = matched$PAM)

```

This annotated phylogeny helps visualize how these *Anolis* species diversified across Cuba, Hispaniola, Jamaica, and Puerto Rico. 

### Partition Species Richness Into Different Components (in situ speciation, import, and export events)
When speciation events were classified above using the `insitu::map_insitu_events()` function, we acknowledged, but did not work with, the number of import and export events. The following functions address all of the different events discussed so far and summarize their relative importance to the system of interest.

First, we will use the `insitu::decompose_diversity()` function with the phylogeny and PAM returned from the `insitu::match_island_phylo()` function, along with the events returned from the `insitu::map_insitu_events()` function to determine the number of species, *in situ* speciation events, import events, and export events associated with each island. 

```{r decompose}
div <- decompose_diversity(phy = matched$phy,
                           events = map_events,
                           PAM = matched$PAM)

# Look at the first 5 rows of the classification data frame
head(div, n = 5)
```

The results of the `insitu::decompose_diversity()` function give us a clearer picture of how species richness accumulated on each function. This diversity decomposition can be further visualized using the `insitu::plot_diversity_decomposition()` function.

```{r plot_div, fig.width = 7, fig.height = 5}
# Visualize diversity decomposition per island in a stacked bar chart
plot_diversity_decomposition(div)
```

The stacked bar plot above accentuates the importance of *in situ* speciation in contributing to species richness for Cuba, Hispaniola, Jamaica, and Puerto Rico, while also highlighting import and export events. We can compare these events statistically through the use of the `insitu::compare_islands()` function, which runs chi-squared and Kruskal-Wallis tests to determine whether event counts, *in situ* rates, and colonization rates differ across islands.

```{r compare_div, warning = FALSE}
# Determine whether:
# - In situ speciation, import, and export counts differ across islands
# - In situ and colonization rates differ across islands
comp_div <- compare_islands(div)
```

The `comp_div` object created using the `insitu::compare_islands()` function is a list of three elements: `counts_test`, `insitu_rate_test`, and `colonization_rate_test`. Each one contains the summary output for each test: 
- `counts_test` is a chi-squared test that determines whether *in situ* speciation, import, and export counts differ across islands.
- `insitu_rate_test` is a Kruskal-Wallis test that determines whether *in situ* speciation rates differ across islands.
- `colonization_rate_test` is a Kruskal-Wallis test that determines whether colonization rates differ across islands.

The results for this example are printed below:

```{r print_compare}
comp_div

```

The p-values for each test indicate that the event counts in general differ between islands, but the *in situ* speciation and colonization rates are not statistically significantly different between islands.

## Rates, Indices, and Support
Up to this point, we have focused on event classification and visualization. In this section, we will cover functionality in the `insitu` package that allows for the calculation of helpful rates and indices that further contextualize the emergence of biodiversity on the islands of interest.

### Determine Colonization Rates
Species richness results from the interplay of speciation, extinction, and dispersal. Determining the rate at which lineages colonize islands helps provide a clearer picture of the drivers of species richness. The `insitu::colonization_rate()` function determines how many colonization events each species is associated with and provides a colonization rate associated with that species' lineage.

```{r col-rates}
# Colonization rate (how many times each species has colonized an island)
colonization_rate(phy = matched$phy, events = map_events)

```

Here, we can see that only *Anolis sagrei* and *Anolis distichus* were involved in colonization events.

### Calculate the *in situ* Speciation Index
In an effort to standardize comparisons across islands to determine the relative importance of *in situ* speciation events in contributing to species richness, the `insitu` package includes a function to calculate the `isci` index. This index represents the number of *in situ* speciation events on an island divided by the total branch length of the island's subtree.

```{r index}
# Per-island in situ speciation index
insitu_speciation_index(phy = matched$phy, events = map_events)
```

In this example, Puerto Rico has the largest `isci` value, indicating that *in situ* speciation contributed more species over a shorter amount of evolutionary time compared to the other islands.

### Determine Node Support
The `insitu::map_insitu_events()` function returns probabilities for the most likely ancestral states among all islands for each node. If the user would like to conduct additional sensitivity analyses related to these ancestral state probabilities, the `insitu` package provides a function that displays the ancestral state probabilities for each island and node combination. It also determines whether the given probability corresponds with low (probability less than 0.5), moderate (probability between 0.5 and 0.75), or high (probability above 0.75) support.

```{r confidence}
# Determine per-node support for in situ speciation events
insitu_conf <- insitu_confidence(recons = asr_insitu)

head(insitu_conf, n = 30)

```

### Incorporating Island Ontogeny
Island ontogeny describes the different stages of emergence and subsistence that oceanic islands experience, and can be thought of as an island's "life cycle." There exists a balance between the age of an island and it species richness: old, large islands often have more species compared to young, small islands, but once an island becomes old enough, it begins to lose area and habitat as it erodes into the ocean. Whittaker et al. (2008) incorporated island ontogeny in their General Dynamic Model of island biogeography, which we suggest users review for additional context.

The `insitu` R package focuses on metrics that help us understand the role *in situ* speciation has in contributing to species richness on island systems. What do we do if an estimated *in situ* speciation event on an internal node is older than the island itself? By incorporating the age of islands in the dataset, we are able to determine whether it is appropriate to mark a node as an *in situ* speciation event on an island, or if the speciation event is more likely to have occurred on a continent before the lineage colonized the islands of interest.

Geologists struggle to define exact ages for the Caribbean Islands, so the island ages written here are either estimates or manufactured ages that help us emphasize the functionality of the `insitu` R package. First, we will create a named vector of island ages for the 6 islands included in our example dataset. Then, we will use this named vector, along with the `map_events` object created by `insitu::map_insitu_events()` above and the matched phylogeny to determine whether the speciation events identified are realistic depending on the age of the islands

```{r ontogeny1}
# Create named numeric vector of island ages
ages <- c(42,
          50,
          25,
          46,
          10,
          15)
island_names <- c("Cuba",
                  "Hispaniola",
                  "Jamaica",
                  "Puerto Rico",
                  "Cat Island",
                  "Ile de la Tortue")
# Assign the names to the numeric vector
names(ages) <- island_names

# Flag in situ speciation events if they are too young
age_correct_events(events = map_events, 
                   phy = matched$phy, 
                   island_ages = ages, 
                   remove_predating = FALSE)

```

Here, we find that two *in situ* speciation events predate the islands on which they are estimated to occur: node 32 associated with Cuba, and node 35 associated with Jamaica. If we set the `remove_predating` parameter to `TRUE`, the function will remove these events and the returned data frame can be used to plot *in situ* speciation events on the phylogeny as above.

```{r ontogeny2}
# Remove in situ speciation events if they are too young
new_events <- age_correct_events(events = map_events, 
                                 phy = matched$phy, 
                                 island_ages = ages, 
                                 remove_predating = TRUE)

# Plot phylogeny with newly filtered events
plot_classified_phylo(phy = matched$phy, events = new_events, PAM = matched$PAM)

```

We can also incorporate island age into the calculation of `isci`, the *in situ* speciation rate index. The `insitu::insitu_rate_by_island_age()` function uses the phylogeny, event object created by `insitu::age_correct_events()`, and the named numeric vector of island ages created above to calculate the corrected `isci` value for each island.

```{r age-index}
insitu_rate_by_island_age(phy = matched$phy, events = new_events, island_ages = ages)
```

## Literature Cited
* Patton, A. H., Harmon, L. J., del Rosario Castañeda, M., Frank, H. K., Donihue, C. M., Herrel, A., & Losos, J. B. (2021). When adaptive radiations collide: Different evolutionary trajectories between and within island and mainland lizard clades. Proceedings of the National Academy of Sciences, 118(42), e2024451118
* Whittaker, R. J., Triantis, K. A., & Ladie, R. J. (2008). A general dynamic theory of oceanic island biogeography. Journal of Biogeography, 35(6), 977–994.
