---
title: "Stochastic character mapping with insitu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stochastic character mapping with insitu}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

`run_geo_asr` estimates a single maximum likelihood ancestral state
reconstruction for each island. Stochastic character mapping goes further
by sampling many possible evolutionary histories consistent with the data,
giving you a posterior distribution of in-situ event classifications rather
than a single point estimate.

`simmap_insitu` wraps `phytools::make.simmap` and returns the same output
structure as `run_geo_asr`, so it plugs directly into `map_insitu_events`
and everything downstream without any changes to the rest of your workflow.
It also accepts a list of trees, propagating phylogenetic uncertainty from
a posterior sample alongside ASR uncertainty from the stochastic maps.

## Setup

Load and prepare the same *Anolis* dataset used in the Getting Started
vignette. Both vignettes use the same objects, so if you are working
through them in order you can skip this step.

```{r setup_data, warning = FALSE}
library(insitu)

tree <- ape::read.tree(
  system.file("extdata/Patton_etal_trimmed.tree", package = "insitu")
)
dat  <- read.csv(
  system.file("extdata/anolis_dat.csv", package = "insitu")
)

tree    <- prep_phylo(tree)
matched <- match_island_phylo(phy = tree, locs = dat)
```

## Running stochastic character mapping

Pass the matched tree and PAM to `simmap_insitu` with a transition model
and the number of stochastic maps to draw. More maps give more stable
posterior averages at the cost of computation time; 100 is reasonable for
publication analyses, 10 is sufficient for exploration.

```{r simmap, warning = FALSE}
sim_results <- simmap_insitu(
  phy   = matched$phy,
  PAM   = matched$PAM,
  model = "ER",
  nsim  = 10
)
```

The result is a named list in exactly the same format as the output of
`run_geo_asr`. The probabilities at each node now reflect the average
across all stochastic maps rather than a single likelihood estimate.

## Propagating phylogenetic uncertainty

If you have a posterior sample of trees from a Bayesian analysis, pass
them as a list. `simmap_insitu` runs the stochastic maps on each tree and
averages the posteriors across both trees and maps. This is the recommended
approach for publication analyses where phylogenetic uncertainty is
non-trivial.

```{r simmap_multi, eval = FALSE}
# trees is a multiPhylo object from a Bayesian posterior sample
sim_results_multi <- simmap_insitu(
  phy   = trees,    # list of phylo objects
  PAM   = matched$PAM,
  model = "ER",
  nsim  = 10
)
```

## Classifying events and comparing to the ML reconstruction

Event classification from `simmap_insitu` works identically to
classification from `run_geo_asr`. Pass the result directly to
`map_insitu_events`.

```{r classify}
events_simmap <- map_insitu_events(
  recons    = sim_results,
  phy       = matched$phy,
  PAM       = matched$PAM,
  threshold = 0.5
)

head(events_simmap, n = 5)
```

Comparing the simmap classification to the ML classification from
`run_geo_asr` is a quick way to assess whether the two approaches
agree. Nodes where they disagree are candidates for closer inspection
using `insitu_confidence`.

```{r compare, warning = FALSE}
asr_ml <- run_geo_asr(phy = matched$phy, PAM = matched$PAM, model = "ER")

events_ml <- map_insitu_events(
  recons    = asr_ml,
  phy       = matched$phy,
  PAM       = matched$PAM,
  threshold = 0.5
)

cat("In-situ events (ML):     ", sum(events_ml$in_situ    == TRUE, na.rm = TRUE), "\n")
cat("In-situ events (simmap): ", sum(events_simmap$in_situ == TRUE, na.rm = TRUE), "\n")
```

## Visualizing the simmap classification

`plot_classified_phylo` works with either reconstruction. Nodes where the
simmap posterior probability is close to the threshold will show visually
in the same way, but a wider ribbon of support can be inspected more
directly with `insitu_confidence`.

```{r plot, fig.width = 9, fig.height = 6}
plot_classified_phylo(phy = matched$phy, events = events_simmap, PAM = matched$PAM)
```

## Checking node-level support

`insitu_confidence` extracts the per-node per-island posterior probabilities
directly from the reconstruction object, whether from `run_geo_asr` or
`simmap_insitu`. Nodes with low support across all islands are the ones most
sensitive to threshold choice.

```{r confidence}
conf <- insitu_confidence(recons = sim_results)

# Nodes with high support for any island
conf[conf$support == "high", ]
```

## Downstream analysis

From this point the workflow is identical to the Getting Started vignette.
Diversity decomposition, the in-situ speciation index, and island ontogeny
correction all accept the events table from either reconstruction approach.

```{r downstream}
div <- decompose_diversity(
  phy    = matched$phy,
  events = events_simmap,
  PAM    = matched$PAM
)

div
```
